---
title: "Lab 05 - Wrangling Spatial Data "
author: "Lilly McClendon"
date: "02-21-2025"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox) 
```

```{r load-data, message = FALSE}
states <- read_csv("data/states.csv")
```

### Exercise 1

```{r view-dennys-laquinta-data}
view(dennys)
view(laquinta)
```

```{r filter-dennys-alaska}
dn_ak <- dennys %>%
  filter(state == "AK")
nrow(dn_ak)
view(dn_ak)
```

```{r filter-laquinta-alaska}
lq_ak <- laquinta %>%
  filter(state == "AK")
nrow(lq_ak)
view(lq_ak)
```

### Exercise 2

#### Total combinations of Denny's and La Quinta Locations 

```{r filter-dennys-laquinta-alaska-combinations}
total_combinations <- nrow(dn_ak) * nrow(lq_ak)
print(total_combinations)
```

There are 6 total combinations of Denny's and La Quinta locations in Alaska. 


### Exercise 3

#### Joining Denny's in Alaska and La Quinta in Alaska datasets 

```{r join-dennys-ak-laquinta-ak}
dn_lq_ak <- full_join(dn_ak, lq_ak, 
                      by = "state")
dn_lq_ak
```

### Exercise 4

```{r observations-variablenames-joined-dn_lq_ak}
library(dplyr)
count(dn_lq_ak)
colnames(dn_lq_ak)
```

#### Number of observations in joined dn_lq_ak data frame

There are 6 observations in the joined dn_lq_ak data frame, which matches up with the number of total combinations (6) we calculated. 

#### Names of variables in joined dn_lq_ak data frame

The names of the variables in the joined dn_lq_ak data frame are address.x, 
city.x, state, zip.x, longitude.x, latitude.x, address.y, city.y, zip.y, longitude.y, and latitude.y. 

### Exercise 5

#### Tidyverse function to add a new variable to data frame while keeping existing variables 

The mutate function in Tidyverse can be used to add a new variable to a data frame while keeping the existing variables. 

```{r Haversine-distance-formula}
haversine <- function(long1, lat1, long2, lat2, round = 3) {
  # convert to radians
  long1 <- long1 * pi / 180
  lat1 <- lat1 * pi / 180
  long2 <- long2 * pi / 180
  lat2 <- lat2 * pi / 180

  R <- 6371 # Earth mean radius in km

  a <- sin((lat2 - lat1) / 2)^2 + cos(lat1) * cos(lat2) * sin((long2 - long1) / 2)^2
  d <- R * 2 * asin(sqrt(a))

  return(round(d, round)) # distance in km
}
```

### Exercise 6

```{r add_distance_to_dn_lq_ak}
dn_lq_ak <- dn_lq_ak %>%
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, round = 3))
```

### Exercise 7 

```{r min-dist-from-each-dennys-laquinta }
dn_lq_ak_mindist <- dn_lq_ak %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))
```

### Exercise 8 

``` {r descriptive-stats}
summary(dn_lq_ak_mindist)
IQR(dn_lq_ak_mindist$closest)
table <- table(dn_lq_ak_mindist)
print(table)
barplot(table)
```

```{r outliers}

IQR_ak <- 1.982 
quart1_ak <- 3.616 
quart3_ak <- 5.598 
print(quart1_ak - (IQR_ak*1.5))
print(quart3_ak + (IQR_ak*1.5))
```

#### Description of the distribution of the distances between a Denny's and La Quinta for each Denny's location in Alaska. 

The IQR method for identifying outliers has determined that none of the three minimum distances between a Denny's and La Quinta are outliers. The distances range from 2.035 to 5. 998, with the average distance equaling 4.410 and a median of 5.197.  

```{r barplot-visualization}
ggplot(dn_lq_ak_mindist, aes(x = forcats::fct_reorder(address.x, desc(closest)), y = closest)) + 
  geom_col() + 
  labs(
    title = "Distance between Denny's and La Quinta Locations in AK",
    x = "Denny's Address",
    y = "Distance (km)"
  )
```

### Exercise 9

```{r filter-dennys-north-carolina}
dn_nc <- dennys %>%
  filter(state == "NC")
nrow(dn_nc)
view(dn_nc)
```

```{r filter-laquinta-north-carolina}
lq_nc <- laquinta %>%
  filter(state == "NC")
nrow(lq_nc)
view(lq_nc)
```

```{r filter-dennys-laquinta-north-carolina-combinations}
total_combinations_nc <- nrow(dn_nc) * nrow(lq_nc)
print(total_combinations_nc)
```

There are 336 total combinations of Denny's and La Quinta locations in Texas. 

```{r join-dennys-nc-laquinta-nc}
dn_lq_nc <- full_join(dn_nc, lq_nc, 
                      by = "state")
dn_lq_nc
```

```{r observations-variablenames-joined-dn_lq_nc}
library(dplyr)
count(dn_lq_nc)
colnames(dn_lq_nc)
```

```{r add_distance_to_dn_lq_nc}
dn_lq_nc <- dn_lq_nc %>%
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, round = 3))
```

```{r min-dist-from-each-dennys-laquinta-nc }
dn_lq_nc_mindist <- dn_lq_nc %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))
```

``` {r descriptive-stats-nc}
summary(dn_lq_nc_mindist)
IQR(dn_lq_nc_mindist$closest)
table <- table(dn_lq_nc_mindist)
```

```{r outliers-nc}

IQR_nc <- 71.597 
quart1_nc <- 22.388
quart3_nc <- 93.985
print(quart1_nc - (IQR_nc*1.5))
print(quart3_nc + (IQR_nc*1.5))
```

#### Description of the distribution of the distances between a Denny's and La Quinta for each Denny's location in North Carolina. 

The IQR method for identifying outliers has determined that none of the 28 minimum distances between a Denny's and La Quinta are outliers in North Carolina. The distances range from 1.779 to 187.935, with the average distance equaling 65.444 and a median of 53.456.  

```{r barplot-visualization-nc}
ggplot(dn_lq_nc_mindist, aes(x = forcats::fct_reorder(address.x, desc(closest)), y = closest)) + 
  geom_col() + 
  labs(
    title = "Distance between Denny's and La Quinta Locations in NC",
    x = "Denny's Address",
    y = "Distance (km)"
  ) + 
  coord_flip()
```

### Exercise 10 

```{r filter-dennys-texas}
dn_tx <- dennys %>%
  filter(state == "TX")
nrow(dn_tx)
view(dn_tx)
```

```{r filter-laquinta-texas}
lq_tx <- laquinta %>%
  filter(state == "TX")
nrow(lq_tx)
view(lq_tx)
```

#### Total combinations of Denny's and La Quinta Locations in Texas 

```{r filter-dennys-laquinta-texas-combinations}
total_combinations_tx <- nrow(dn_tx) * nrow(lq_tx)
print(total_combinations_tx)
```

There are 47,400 total combinations of Denny's and La Quinta locations in Texas. 

#### Joining Denny's in Texas and La Quinta in Texas datasets 
```{r join-dennys-tx-laquinta-tx}
dn_lq_tx <- full_join(dn_tx, lq_tx, 
                      by = "state")
dn_lq_tx
```

```{r observations-variablenames-joined-dn_lq_tx}
library(dplyr)
count(dn_lq_tx)
colnames(dn_lq_tx)
```
#### Number of observations in joined dn_lq_tx data frame

There are 47,400 observations in the joined dn_lq_ak data frame, which matches up with the number of total combinations (47,400) we calculated. 

#### Names of variables in joined dn_lq_tx data frame

The names of the variables in the joined dn_lq_tx data frame are address.x, city.x, state, zip.x, longitude.x, latitude.x, address.y, city.y, zip.y, longitude.y, and latitude.y. 


```{r add_distance_to_dn_lq_tx}
dn_lq_tx <- dn_lq_tx %>%
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, round = 3))
```

```{r min-dist-from-each-dennys-laquinta-tx }
dn_lq_tx_mindist <- dn_lq_tx %>%
  group_by(address.x) %>%
  summarize(closest = min(distance)) 
dn_lq_tx_mindist[order(dn_lq_tx_mindist$closest, decreasing= TRUE),]
```

``` {r descriptive-stats-tx}
summary(dn_lq_tx_mindist)
IQR(dn_lq_tx_mindist$closest)
table <- table(dn_lq_tx_mindist)
```

```{r outliers-tx}

IQR_tx <- 5.89975
quart1_tx <- 0.7305 
quart3_tx <- 6.6303 
print(quart1_tx - (IQR_tx*1.5))
print(quart3_tx + (IQR_tx*1.5))
```

#### Description of the distribution of the distances between a Denny's and La Quinta for each Denny's location. 

The IQR method for identifying outliers has determined that 14 of the 200 minimum distances between a Denny's and La Quinta are outliers in Texas. The distances range from 0.0160 to 60.5820 with the average distance equaling 5.7918 and a median of 3.3715.  

```{r barplot-visualization-texas}
ggplot(dn_lq_tx_mindist, aes(x = forcats::fct_reorder(address.x, desc(closest)), y = closest)) + 
  geom_col() + 
  labs(
    title = "Distance between Denny's and La Quinta Locations in TX",
    x = "Denny's Address",
    y = "Distance (km)"
  ) +
coord_flip()
```

### Exercise 11 

```{r filter-dennys-newjersey}
dn_nj <- dennys %>%
  filter(state == "NJ")
nrow(dn_nj)
view(dn_nj)
```

```{r filter-laquinta-nj}
lq_nj <- laquinta %>%
  filter(state == "NJ")
nrow(lq_nj)
view(lq_nj)
```

```{r filter-dennys-laquinta-newjersey-combinations}
total_combinations_nj <- nrow(dn_nj) * nrow(lq_nj)
print(total_combinations_nj)
```

```{r join-dennys-nj-laquinta-nj}
dn_lq_nj <- full_join(dn_nj, lq_nj, 
                      by = "state")
dn_lq_nj
```

```{r observations-variablenames-joined-dn_lq_nj}
library(dplyr)
count(dn_lq_nj)
colnames(dn_lq_nj)
```

```{r add_distance_to_dn_lq_nj}
dn_lq_nj <- dn_lq_nj %>%
  mutate(distance = haversine(longitude.x, latitude.x, longitude.y, latitude.y, round = 3))
```

```{r min-dist-from-each-dennys-laquinta_nj}
dn_lq_nj_mindist <- dn_lq_nj %>%
  group_by(address.x) %>%
  summarize(closest = min(distance))
```

``` {r descriptive-stats-nj}
summary(dn_lq_nj_mindist)
IQR(dn_lq_nj_mindist$closest)
table <- table(dn_lq_nj_mindist)
```

```{r outliers-nj}

IQR_nj <- 23.524
quart1_nj <- 30.88 
quart3_nj <- 54.40
print(quart1_nj - (IQR_nj*1.5))
print(quart3_nj + (IQR_nj*1.5))
```

#### Description of the distribution of the distances between a Denny's and La Quinta for each Denny's location in New Jersey 

The IQR method for identifying outliers has determined that none of the 10 minimum distances between a Denny's and La Quinta are outliers. The distances range from 15.39 to 69.13, with the average distance equaling 42.29 and a median of 41.52.  

```{r barplot-visualization-nj}
ggplot(dn_lq_nj_mindist, aes(x = forcats::fct_reorder(address.x, desc(closest)), y = closest)) + 
  geom_col() + 
  labs(
    title = "Distance between Denny's and La Quinta Locations in NJ",
    x = "Denny's Address",
    y = "Distance (km)"
  ) +
coord_flip()
```

### Exercise 12 

Among the states that I looked at (Alaska, North Carolina, Texas, and New Jersey) Mitch Hedberg's joke is most likely to hold true is Alaska. Among the minimum distances, the greatest distance was approximately 6km. Whereas in the when North Carolina, Texas and New Jersey were examined, the among the minimum distances, the greatest distances were 187, 60.58, and 69.13 respectively. Although Texas has a smaller median distance value (3.37) than Alaska (5.20), I would still assert that Alaska is the state where Mitch Hedberg's joke is most likely to hold true. I think this becuase of the maximum distance in Alaska of approximately 6 kilometers. In the right conditions, a person can see roughly 5 kilometers. Therefore, in Alaska, a person would be able to literally see a la quinta from every Denny's location which would make the joke seem the most plausible. In texas, there is a much larger range of values so there would be many occasions where there is a Denny's without a la quinta close nearby. 